Life in the canopy: Phylogenetic filtering and morphological canalization in epiphytic ferns

### 2018-03-21 manuscript.Rnw
Change hard-coded results values to Sexpr for almost all results and discussion
Add mean_trait_values to analyses. Still need to fix sla for Selliguea plantaginea

<<run_analyses, results="hide">>=
# set working directory, run all analyses, save results objects
setwd(here::here())
source(file = "bin/microclimate.R")
source(file = "bin/trait_PCA.R")
source(file = "bin/cont_phylosig.R")
source(file = "bin/binary_phylosig.R")
source(file = "bin/sporo_PIC.R")
source(file = "bin/correlated_evo.R")
source(file = "bin/binaryPGLMM.R")
source(file = "bin/functional_diversity.R")
save(list = ls(all = TRUE), file= "results.RData")
@

<<get_result_func, echo=FALSE>>=
# simple function to get rounded results with less typing
get_result <- function (results_table, col_name, row_name, signif=FALSE, digits=3) {
  if (signif == FALSE) {
    value <- round(results_table[row_name, col_name], digits=3)
  } else if (signif == TRUE) {
    value <- signif(results_table[row_name, col_name], digits=digits)
  }
  return(value)
}
@

@@

Most sporophyte traits, including all those related to plant size, are significantly correlated with growth habit while taking phylogeny into account (\tab \ref{table:PIC_sporo}). Epiphytic species tend to have smaller values for all traits (\tab \ref{table:PIC_sporo}; contrasts were made by subtracting epiphytic from terrestrial values). Short stipe length in particular appears to be strongly correlated with epiphytic growth (phylogenetically independent contrasts, $\pval = \Sexpr{get_result(PIC.results, "pval", "stipe", digits=2, signif=TRUE)}$), as well as frond width ($\pval = \Sexpr{get_result(PIC.results, "pval", "width", digits=2, signif=TRUE)}$) and degree of laminae dissection ($\pval = \Sexpr{get_result(PIC.results, "pval", "dissection", digits=2, signif=TRUE)}$). For the gametophyte traits, only morphotype (non-cordate vs. cordate) was significantly correlated with growth habit (Pagel's test of correlated evolution, $\pval = 0.0011$; \tab \ref{table:pagel_gameto}). Quantitatively similar results were obtained using PGLMMs, except that no significant relationship was observed between rhizome diameter and growth habit, whereas SLA did show a significant relationship with growth habit (\tabs \ref{supp_table:PGLMM_sporo}, \ref{supp_table:PGLMM_gameto}).

@@

On a single-trait basis, distribution of community-weighted mean values (CWMs) showed significant correlation with elevation for some traits, but no significant relationship was detected for any traits with minimum RH (\fig \ref{fig:CWM}A). SLA decreased with elevation for epiphytic (linear model, $\rval^2 = 0.6787$, $\pval <0.001$) and terrestrial (linear model, $\rval^2 = 0.8118$, $\pval <0.001$) communities. Degree of lamina dissection decreased with elevation in epiphytes (linear model, $\rval^2 = 0.3634$, $\pval = 0.0104$), but not terrestrial communities. Stipe length showed a very weak, but significant, increase with elevation for epiphytic communities (linear model, $\rval^2 = 0.2721$, $\pval = 0.0317$). Rhizome diameter decreased with elevation for terrestrial communities (linear model, $\rval^2 = 0.238$, $\pval = 0.047$), but increased slightly in epiphytes (linear model, $\rval^2 = 0.3525$, $\pval = 0.0120$).

We observed a total of 104 species across all plots, including 48 epiphytic and 56 terrestrial species. Mean species richness was $11.12 \pm 4.6$ \si{\species} per plot for epiphytes and $12.35 \pm 5.0$ \si{\species} per plot for terrestrial species (not significantly different; one sided \tval-test, $\pval = 0.23$; \fig \ref{fig:FDIV}A). Species richness increased linearly with elevation and minimum RH for epiphytes, but not terrestrial communities (\fig \ref{fig:FDIV}B). Trait richness (FRic) was lower in epiphytic vs. terrestrial communities overall (one sided \tval-test, $\pval < 0.001$), whereas functional evenness (FEve) and diversity (FDiv) did not differ between the two. FEve decreased with elevation, but not minimum RH, in terrestrial communities (\fig \ref{fig:FDIV}A). Epiphytic communities showed a trend of increasing trait richness along the environmental gradient, with greater FRic at high elevations (linear model, $\rval^2 = 0.5836$, $\pval < 0.001$) and plots with greater minimum RH (linear model, $\rval^2 = 0.6626$, $\pval = 0.009$). Epiphytic communities were more phylogenetically clustered than terrestrial communities overall when measured using MNTD (one sided \tval-test, $\pval < 0.001$) but not MPD ($\pval = 0.30$; \fig \ref{fig:FDIV}B). No significant trend was detected in phylogenetic community structure with environment in either terrestrial or epiphytic communities (\fig \ref{fig:FDIV}A).

@@

Surprisingly, SLA was only slightly lower in epiphytes relative to terrestrial species (\tab\ref{table:PIC_sporo}, \fig\ref{fig:CWM}B). SLA is an important part of the leaf economic spectrum, which posits that leaf traits including SLA, N and C content, and leaf life span vary along well-defined axes such that plants generally fall into one of two categories: fast-growing species with high SLA, low nutrient content, and short life spans vs. slow-growing species with low SLA, high nutrient content, and long life spans \citep{Wright2004a}. We hypothesized that epiphytes would have lower SLA, given their more extreme growing conditions. Indeed, the species with the lowest SLA values were epiphytes (\eg, \textit{Selliguea plantaginea} Brack., \SI{8.7}{\square\meter\per\kilo\gram}; \textit{Davallia anderssonii} Mett. ex Kuhn, \SI{8.43}{\square\meter\per\kilo\gram}), but this relationship was not significant across the dataset (\tab\ref{table:PIC_sporo}, \fig\ref{fig:CWM}). The tradeoffs posited in the leaf economics spectrum have been most thoroughly documented in global studies including the widest possible range of taxa and environments, and do not always hold up in local studies \citep{Petter2016}. For example, \citet{Zhu2016} did not find a significant relationship between SLA and leaf life span in a survey of 16 ferns across a disturbance (light) gradient in China. It is possible that other small-scale effects (\eg, microclimate) obscure this pattern in local studies such as ours and that of \citet{Zhu2016}. This is supported by the strong, parallel relationship in CWM values of SLA with elevation in both epiphytes and terrestrial species (\fig\ref{fig:CWM}A).

@@

We observed the predicted trend of decreased trait richness (FRic) in epiphytic relative to terrestrial communities overall, and a significant positive correlation of FRic with humidity in epiphytes (\ie, decreased FRic at drier, low elevation plots and increased FRic at more humid, high elevation plots; \fig\ref{fig:FDIV}A). However, we did not observe similar patterns in the other two measures of functional diversity, evenness (FEve) and divergence (FDiv) (\fig\ref{fig:FDIV}B). Indeed, FEve decreased with elevation (but not minimum RH) in terrestrial communities (\fig\ref{fig:FDIV}A). Overall, the lack of a clear pattern in functional evenness and divergence suggests that within each trait pool, there is no significant difference in trait spacing between epiphytes and terrestrial species. This could be maintained by niche partitioning within epiphytic and terrestrial habitats. Epiphytic habitats are highly heterogeneous, and microclimatic conditions vary with height along the host and distance from the trunk \citep{Hietz1995}. The vertical stratification of epiphytic species into different zones along height gradients in host trees is well documented \citep{Zotz2007}. In a study comparing terrestrial and epiphytic ferns along an elevational gradient from \SIrange{30}{3000}{\meter} in Costa Rica, \citet{WatkinsJr.2006a} found that epiphytes tended to sort into low-trunk and canopy zones, with only 18\% found in both, and no species that overlapped between canopy and terrestrial habitats. It is possible that epiphytic growth per se limits overall morphological variation in epiphytic ferns, but that species with different traits are then sorted into different niches.

>>>>

<<run_analyses, results="hide">>=
# set working directory, run all analyses, save results objects
setwd(here::here())
source(file = "bin/microclimate.R")
source(file = "bin/trait_PCA.R")
source(file = "bin/cont_phylosig.R")
source(file = "bin/binary_phylosig.R")
source(file = "bin/sporo_PIC.R")
source(file = "bin/correlated_evo.R")
source(file = "bin/binaryPGLMM.R")
source(file = "bin/functional_diversity.R")
source(file = "bin/mean_trait_values.R") # this must be last to keep sla.grand.mean
save(list = ls(all = TRUE), file= "results.RData")
@

<<get_result_func>>=
# simple function to get rounded results with less typing
library(dplyr) # use to filter df prior to get_result in some cases
get_result <- function (results_table, col_name, row_name=1, signif=FALSE, digits=3) {
  if (signif == FALSE) {
    value <- round(results_table[row_name, col_name], digits=3)
  } else if (signif == TRUE) {
    value <- signif(results_table[row_name, col_name], digits=digits)
  }
  return(value)
}

@

@@

Most sporophyte traits, including all those related to plant size, are significantly correlated with growth habit while taking phylogeny into account (\tab \ref{table:PIC_sporo}). Epiphytic species tend to have smaller values for all traits (\tab \ref{table:PIC_sporo}; contrasts were made by subtracting epiphytic from terrestrial values). Short stipe length in particular appears to be strongly correlated with epiphytic growth (phylogenetically independent contrasts, $\pval = \Sexpr{get_result(PIC.results, "pval", "stipe", digits=2, signif=TRUE)}$), as well as frond width ($\pval = \Sexpr{get_result(PIC.results, "pval", "width", digits=2, signif=TRUE)}$) and degree of laminae dissection ($\pval = \Sexpr{get_result(PIC.results, "pval", "dissection", digits=2, signif=TRUE)}$). For the gametophyte traits, only morphotype (non-cordate vs. cordate) was significantly correlated with growth habit (Pagel's test of correlated evolution, $\pval = \Sexpr{get_result(pagel.results, "pval", "morph_binary", digits=2, signif=TRUE)}$; \tab \ref{table:pagel_gameto}). Quantitatively similar results were obtained using PGLMMs, except that no significant relationship was observed between rhizome diameter and growth habit, whereas SLA did show a significant relationship with growth habit (\tabs \ref{supp_table:PGLMM_sporo}, \ref{supp_table:PGLMM_gameto}).

@@

<<calculate_richness>>=
### calculate number of species observed in plots total and by growth habit
# use only sites on Moorea
sites <- mooreaferns::sites[grep("Aorai", mooreaferns::sites$site, invert=TRUE), ]
# load untransformed traits
traits <- mooreaferns::fern_traits
# load community data
comm <- mooreaferns::sporocomm
# trim community data to only moorea sites
comm <- comm[rownames(comm) %in% sites$site, ]
# keep only species with at least one observation
comm <- comm[, colSums(comm) > 0]
# trim trait data to only species with at least one observation
traits <- traits[traits$species %in% colnames(comm), ]
@

On a single-trait basis, distribution of community-weighted mean values (CWMs) showed significant correlation with elevation for some traits, but no significant relationship was detected for any traits with minimum RH (\fig \ref{fig:CWM}A). SLA decreased with elevation for epiphytic (linear model, $\rval^2 = \Sexpr{get_result(filter(univar.el.results, variable=="sla", habit=="epiphytic"), "r.squared.plain")}$, $\pval = \Sexpr{get_result(filter(univar.el.results, variable=="sla", habit=="epiphytic"), "p", signif=TRUE)}$) and terrestrial (linear model, $\rval^2 = \Sexpr{get_result(filter(univar.el.results, variable=="sla", habit=="terrestrial"), "r.squared.plain")}$, $\pval = \Sexpr{get_result(filter(univar.el.results, variable=="sla", habit=="terrestrial"), "p", signif=TRUE)}$) communities. Degree of lamina dissection decreased with elevation in epiphytes (linear model, $\rval^2 = \Sexpr{get_result(filter(univar.el.results, variable=="dissection", habit=="epiphytic"), "r.squared.plain")}$, $\pval = \Sexpr{get_result(filter(univar.el.results, variable=="dissection", habit=="epiphytic"), "p", signif=TRUE)}$), but not terrestrial communities. Stipe length showed a small, but significant, increase with elevation for epiphytic communities (linear model, $\rval^2 = \Sexpr{get_result(filter(univar.el.results, variable=="stipe", habit=="epiphytic"), "r.squared.plain")}$, $\pval = \Sexpr{get_result(filter(univar.el.results, variable=="stipe", habit=="epiphytic"), "p", signif=TRUE)}$). Rhizome diameter decreased with elevation for terrestrial communities (linear model, $\rval^2 = \Sexpr{get_result(filter(univar.el.results, variable=="rhizome", habit=="terrestrial"), "r.squared.plain")}$, $\pval = \Sexpr{get_result(filter(univar.el.results, variable=="rhizome", habit=="terrestrial"), "p", signif=TRUE)}$), but increased slightly in epiphytes (linear model, $\rval^2 = \Sexpr{get_result(filter(univar.el.results, variable=="rhizome", habit=="epiphytic"), "r.squared.plain")}$, $\pval = \Sexpr{get_result(filter(univar.el.results, variable=="rhizome", habit=="epiphytic"), "p", signif=TRUE)}$).

We observed a total of \Sexpr{ncol(comm)} species across all plots, including \Sexpr{length(traits$habit[traits$habit == "epiphytic"])} epiphytic and \Sexpr{length(traits$habit[traits$habit == "terrestrial"])} terrestrial species. Mean species richness was $\Sexpr{get_result(filter(multvar.t.results, resp_var=="ntaxa", habit=="epiphytic"), "mean", signif=TRUE)} \pm \stdev \Sexpr{get_result(filter(multvar.t.results, resp_var=="ntaxa", habit=="epiphytic"), "sd", signif=TRUE)}$ \si{\species} per plot for epiphytes and $\Sexpr{get_result(filter(multvar.t.results, resp_var=="ntaxa", habit=="terrestrial"), "mean", signif=TRUE)} \pm \stdev \Sexpr{get_result(filter(multvar.t.results, resp_var=="ntaxa", habit=="terrestrial"), "sd", signif=TRUE)}$ \si{\species} per plot for terrestrial species (not significantly different; two-sided \tval-test, $\pval = \Sexpr{get_result(filter(multvar.t.results, resp_var=="ntaxa", habit=="terrestrial"), "pval", signif=TRUE)}$; \fig \ref{fig:FDIV}A). Species richness increased linearly with elevation and minimum RH for epiphytes, but not terrestrial communities (\fig \ref{fig:FDIV}B). Trait richness (FRic) was lower in epiphytic vs. terrestrial communities (two-sided \tval-test, $\pval = \Sexpr{get_result(filter(multvar.t.results, resp_var=="FRic", habit=="terrestrial"), "pval", signif=TRUE)}$), whereas functional diversity (FDiv) was lower in terrestrial communities (two-sided \tval-test, $\pval = \Sexpr{get_result(filter(multvar.t.results, resp_var=="FDiv", habit=="terrestrial"), "pval", signif=TRUE)}$), and functional evenness (FEve) did not differ between the two. FEve decreased with elevation, but not minimum RH, in terrestrial communities (\fig \ref{fig:FDIV}A). Epiphytic communities showed a trend of increasing trait richness along the environmental gradient, with greater FRic at high elevations (linear model, $\rval^2 = \Sexpr{get_result(filter(multvar.el.results, habit=="epiphytic", variable=="FRic"), "r.squared.plain", signif=TRUE)}$, $\pval = \Sexpr{get_result(filter(multvar.el.results, habit=="epiphytic", variable=="FRic"), "p", signif=TRUE, digits=1)}$) and plots with greater minimum RH (linear model, $\rval^2 = \Sexpr{get_result(filter(multvar.env.results, habit=="epiphytic", variable=="FRic"), "r.squared.plain")}$, $\pval = \Sexpr{get_result(filter(multvar.env.results, habit=="epiphytic", variable=="FRic"), "p")}$). Epiphytic communities were more phylogenetically clustered than terrestrial communities overall when measured using MNTD (two-sided \tval-test, $\pval = \Sexpr{get_result(filter(multvar.t.results, habit=="terrestrial", resp_var=="mntd.obs.z"), "pval", signif=TRUE)}$) but not MPD ($\pval = \Sexpr{get_result(filter(multvar.t.results, habit=="terrestrial", resp_var=="mpd.obs.z"), "pval", signif=TRUE, digits=2)}$; \fig \ref{fig:FDIV}B). No significant trend was detected in phylogenetic community structure with environment in either terrestrial or epiphytic communities (\fig \ref{fig:FDIV}A).

@@

Surprisingly, SLA was only slightly lower in epiphytes relative to terrestrial species (\tab\ref{table:PIC_sporo}, \fig\ref{fig:CWM}B). SLA is an important part of the leaf economic spectrum, which posits that leaf traits including SLA, N and C content, and leaf life span vary along well-defined axes such that plants generally fall into one of two categories: fast-growing species with high SLA, low nutrient content, and short life spans vs. slow-growing species with low SLA, high nutrient content, and long life spans \citep{Wright2004a}. We hypothesized that epiphytes would have lower SLA, given their more extreme growing conditions. Indeed, the species with the lowest SLA values were epiphytes (\eg, \textit{Selliguea plantaginea} Brack., \SI{\Sexpr{as.numeric(get_result(filter(sla.grand.mean, species=="Selliguea_plantaginea"), "mean", signif=TRUE))}}{\square\meter\per\kilo\gram}; \textit{Ctenopterella blechnoides} (Grev.) Parris, \SI{\Sexpr{as.numeric(get_result(filter(sla.grand.mean, species=="Ctenopterella_blechnoides"), "mean", signif=TRUE))}}{\square\meter\per\kilo\gram}), but this relationship was not significant across the dataset (\tab\ref{table:PIC_sporo}, \fig\ref{fig:CWM}). The tradeoffs posited in the leaf economics spectrum have been most thoroughly documented in global studies including the widest possible range of taxa and environments, and do not always hold up in local studies \citep{Petter2016}. For example, \citet{Zhu2016} did not find a significant relationship between SLA and leaf life span in a survey of 16 ferns across a disturbance (light) gradient in China. It is possible that other small-scale effects (\eg, microclimate) obscure this pattern in local studies such as ours and that of \citet{Zhu2016}. This is supported by the strong, parallel relationship in CWM values of SLA with elevation in both epiphytes and terrestrial species (\fig\ref{fig:CWM}A).

@@

Results for functional trait richness (FRic) supported our hypotheses: we observed lower FRic in epiphytic relative to terrestrial communities overall, and a significant positive correlation of FRic with humidity in epiphytes (\ie, decreased FRic at drier, low elevation plots and increased FRic at more humid, high elevation plots; \fig\ref{fig:FDIV}A). However, results for the other two measures of functional diversity, evenness (FEve) and divergence (FDiv) (\fig\ref{fig:FDIV}B) did not match our predictions. Rather, FEve decreased with elevation (but not minimum RH) in terrestrial communities, and FDiv was lower in terrestrial communities overall while showing no trend along the gradient (\fig\ref{fig:FDIV}A). The lack of a clear pattern in functional evenness and divergence suggests that within each trait pool, there is no significant difference in trait spacing between epiphytes and terrestrial species. This could be maintained by niche partitioning within epiphytic and terrestrial habitats. Epiphytic habitats are highly heterogeneous, and microclimatic conditions vary with height along the host and distance from the trunk \citep{Hietz1995}. The vertical stratification of epiphytic species into different zones along height gradients in host trees is well documented \citep{Zotz2007}. In a study comparing terrestrial and epiphytic ferns along an elevational gradient from \SIrange{30}{3000}{\meter} in Costa Rica, \citet{WatkinsJr.2006a} found that epiphytes tended to sort into low-trunk and canopy zones, with only 18\% found in both, and no species that overlapped between canopy and terrestrial habitats. It is possible that epiphytic growth per se limits overall morphological variation in epiphytic ferns, but that species with different traits are then sorted into different niches.
